Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems

### 2024-08-17 slides/2024_8_30.tex
Updated slides

\newdate{date}{20}{08}{2024}
\date{\displaydate{date}}
\title[Fairness]{Ensuring Fairness with Transparent Auditing of
Quantitative Bias in AI Systems}
\author[Rex]{
    \begin{CJK}{UTF8}{bsmi}袁至誠\end{CJK}\newline
    Chih-cheng Rex Yuan\newline
    \href{https://rexyuan.com/}{rexyuan.com}
    }
\institute[IIS,AS]{Institute of Information Science, Academia Sinica}

@@

\begin{frame}
    \frametitle{Bias}
    \begin{itemize}
        \item AI is widely used in decision making:
        \begin{itemize}
            \item School admission.
            \item Loan approval.
            \item Hiring.
            \item Policing.
            \item Censorship.
            \item etc..
        \end{itemize}
        \item Decision making by AI may be biased.
        \item Bias can come from several sources:
        \begin{itemize}
            \item Biased data. ML is deisgned to replicate this.
            \item Missing data. The datasets might not be representative.
            \item Biased algorithms. The objective functions might introduce bias.
            \item Sensitive attributes: Age, Gender, ..., etc..
        \end{itemize}
    \end{itemize}
\end{frame}

@@

    Let's set $\epsilon = 0.2$.
    If for some job opening there are $10$ female applicants and
    $100$ male applicants, and there are $8$ accepted females
    and $90$ accepted males:
    \[
        \abs{8/10 - 90/100} = 0.1 < \epsilon \quad\text{so this is fair}
    \]
    while if there were $1$ accepted females and $50$ accepted males:
    \[
        \abs{1/10 - 50/100} = 0.4 > \epsilon \quad\text{so this is unfair}
    \]
\end{frame}

@@

\begin{frame}
    \frametitle{Other Measures}
    \begin{itemize}
        \item Overall accuracy equality:
        \[
            \abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon
        \]
        where $Y = \hat{Y}$ means that the prediction was correct.
        \item Predictive parity:
        \[
            \abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon
        \]
        This requires that the ``positive predictive values'' are similar across
        groups, meaning the probability of an individual with a positive prediction
        actually experiencing a positive outcome.
    \end{itemize}
\end{frame}

@@

\begin{frame}
    \frametitle{Abstraction}
    Our framework tool abstracts the fairness measures:
    \begin{itemize}
        \item A \emph{row} is a lookup table or dictionary.
        $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
        \item A \emph{privileged predicate} $\priv$ takes a row and determines if
        it belongs to the privileged group.
        For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$
        means the privileged group is those with race being Caucasian.
        \item A \emph{positive predicate} $\pos$ takes a row and determines if its
        prediction is positive. For example,
        $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a
        row's prediction is positive if its score is greater than 7.
        \item A \emph{ground truth predicate} $\tru$ takes a row and gives the ground
        truth of the result. For example,
        $\tru(r_i) := r_i(\text{``recid''})$ indicates a
        row's actual recidivism.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Definition}
    We can use these abstractions to define fairness measures.
    For equal opportunity, recall its formal definition:
    \begin{align*}
        \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]}
        \leq \epsilon
    \end{align*}
    To model $Y$, $\hat{Y}$, and $S$, we define the corresponding
    $\tru$, $\pos$, and $\priv$.
    \begin{itemize}
        \item $Y = 1$ if and only if $\tru$ is true
        \item $\hat{Y} = 1$ if and only if $\pos$ is true
        \item $S = 1$ if and only if $\priv$ is true
    \end{itemize}
    This way, we can calculate equal opportunity as a function:
    \begin{align*}
        \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
    \end{align*}
\end{frame}

\begin{frame}
    \frametitle{Application}
    To demonstrate an application, we analyzed the COMPAS dataset.
    We set the predicates:
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
    \end{align*}
    These predicates can be published for transparency.
\end{frame}

@@

>>>>

\newdate{date}{30}{08}{2024}
\date{\displaydate{date}}
\title[Fairness]{Ensuring Fairness with Transparent Auditing of
Quantitative Bias in AI Systems}
\author[Rex]{
    \begin{CJK}{UTF8}{bsmi}袁至誠\end{CJK}\newline
    Chih-cheng Rex Yuan\newline
    \href{https://rexyuan.com/}{rexyuan.com}
    }
\institute[IIS,AS]{Institute of Information Science, Academia Sinica}

@@

\begin{frame}
    \frametitle{AI Bias}
    \begin{itemize}
        \item AI is widely used in decision making:
        \begin{itemize}
            \item School admission.
            \item Loan approval.
            \item Hiring.
            \item Policing.
            \item Censorship.
            \item etc..
        \end{itemize}
        \item Decision making by AI may be biased.
        \item Bias can come from several sources:
        \begin{itemize}
            \item Biased data. ML is deisgned to replicate this.
            \item Missing data. The datasets might not be representative.
            \item Biased algorithms. The objective functions might introduce bias.
            \item Sensitive attributes: Age, Gender, ..., etc..
        \end{itemize}
    \end{itemize}
\end{frame}

@@

    Let's set $\epsilon = 0.2$.
    If for some job opening, male is the privileged group and
    there are $10$ female applicants and
    $100$ male applicants, and there are $8$ accepted females
    and $90$ accepted males:
    \[
        \abs{8/10 - 90/100} = 0.1 < \epsilon \quad\text{so this is fair}
    \]
    while if there were $1$ accepted females and $50$ accepted males:
    \[
        \abs{1/10 - 50/100} = 0.4 > \epsilon \quad\text{so this is unfair}
    \]
\end{frame}

@@

\begin{frame}
    \frametitle{Other Measures}
        \begin{table}[h]
            \centering
            \resizebox{0.93\textwidth}{!}{
            % \resizebox{\columnwidth}{!}{
            \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
                \hline
                \textbf{Fairness Measure} & \textbf{Definition} \\
                \hline
        Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
        \hline
        Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
        \hline
        Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
        \hline
        Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
        \hline
        Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
        \hline
        \multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
        \cline{2-2}
                                        & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
        \hline
        Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
        \hline
        Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
        \hline
        \multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
        \cline{2-2}
                                         & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
        \hline
        Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
        \hline
        Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
        \hline
        Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
        \hline
        Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
        \hline
            \end{tabular}
            }
            \caption{Fairness measures.}
            \label{tab:measures}
        \end{table}
\end{frame}

@@

\begin{frame}
    \frametitle{Predicate}
    We first introduce the idea of a predicate.
    \begin{itemize}
        \item A \emph{predicate} is a decider that determines
        if some condition holds for some given input.
        \item Formally, it is a function that takes
        some type $\alpha$ and returns True or False.
        \item For example, an ``$\text{adult}(x)$'' predicate might
        take a person $x$'s data and returns if they are over 18
        year old.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Abstraction}
    Our framework tool abstracts the fairness measures:
    \begin{itemize}
        \item A \emph{row} is a lookup table or dictionary.
        $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
        \item A \emph{privileged predicate} $\priv$ takes a row and determines if
        it belongs to the privileged group.
        For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$
        means the privileged group is those with race being Caucasian.
        \item A \emph{positive predicate} $\pos$ takes a row and determines if its
        prediction is positive. For example,
        $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a
        row's prediction is positive if its score is greater than 7.
        \item A \emph{ground truth predicate} $\tru$ takes a row and gives the ground
        truth of the result. For example,
        $\tru(r_i) := r_i(\text{``recid''}) == \text{True}$ indicates a
        row's actual recidivism.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Definition}
    We can use these abstractions to define fairness measures.
    For equal opportunity, recall its formal definition:
    \begin{align*}
        \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]}
        \leq \epsilon
    \end{align*}
    To model $Y$, $\hat{Y}$, and $S$, we define the corresponding
    $\tru$, $\pos$, and $\priv$.
    \begin{itemize}
        \item $Y = 1$ if and only if $\tru$ is true
        \item $\hat{Y} = 1$ if and only if $\pos$ is true
        \item $S = 1$ if and only if $\priv$ is true
        \item and vice versa
    \end{itemize}
    This way, we can calculate equal opportunity as a function:
    \begin{align*}
        \text{equal\_opportunity} (\epsilon, \priv, \pos, \tru)
    \end{align*}
\end{frame}

\begin{frame}
    \frametitle{Application}
    For demonstration, we analyzed the COMPAS dataset.
    Let's set the predicates:
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
    \end{align*}
    These predicates can be published for transparency.
\end{frame}

@@

\begin{frame}
    \frametitle{Application}
    \centering
    \includegraphics[width=0.73\textwidth]{Asian}
    \captionof{figure}{Unprivileged Group: Asian (32 rows)}
\end{frame}
